Can Cultural Transmission Explain The Evolution of Cooperative Behaviour

### 2020-03-04 ms/main.tex
revise model

\documentclass{article}
\usepackage[margin=0.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage[displaymath, mathlines]{lineno}
\usepackage{graphicx}
\usepackage[numbers]{natbib}

@@

\textbf{Kin selection}, proposing that natural selection can favour cooperative behaviour between kin. The importance of relatedness to the evolution of cooperation and altruism was shown by Hamilton~\cite{hamilton1964genetical}. According to Hamilton, kin selection causes allele to increase in frequency when the genetic relatedness $r$ of a recipient to an actor multiplied by the benefit $b$ to the recipient is greater than the reproductive cost $c$ to the actor. This is also known as Hamilton's Rule:
\begin{equation} \label{hamiltonrule}
c<rb
\end{equation}
There is an ongoing debate about to what extent kin selection explains evolution of cooperation and altruism.
It has been suggested that kin selection to explain the cooperative behaviour of eusocial insects like the honey bee.
The most significant argument against kin selection is that cooperation can evolve with zero relatedness~\cite{wilson2005kin}. This makes Hamilton's rule incomplete according to Wilson~\cite{wilson2005kin}. Foster et~al.~\cite{foster2006kin} reject this claim. They argue that altruism without relatedness can not evolve. They refer us to Hamilton who claimed that relatedness can arise without recent common ancestry. 
Wilson also criticises kin selection on the grounds that environmental or ecological factors probably be more important than relatedness in determining social actions. On the other hand, Foster et~al.~\cite{foster2006kin} argue that kin selection does not ignore ecology. Hamilton’s rule shows
that environmental factors causing a high benefit-to-cost ratio will favour cooperation.

\textbf{Reciprocity} suggests repeating interactions or individual recognition as key factors for explaining the evolution of cooperation. In \textit{direct reciprocity} there are a repeated encounters between the same two individuals. In every encounter, each player has a choice between cooperation and defection. If I cooperate now, you may cooperate later. Hence, it may pay off to cooperate.
This game-theoretic framework is known as the repeated Prisoner's Dilemma. 
%A well known strategy to play this game is called  'tit-for-tat'. This strategy always starts with a cooperation, then it does whatever the other player has done in the previous round: a cooperation for a cooperation, a defection for a defection. There a unlimited number of possible strategies to play this game. However, 
Direct reciprocity can only lead to the evolution of cooperation if the cost is smaller than $w$ the probability for another encounter between the same two individuals multiplied by the benefit. 
\begin{equation} \label{reciprocity}
c<bw
\end{equation}
Direct reciprocity assumes that both players are in a position to cooperate. Direct reciprocity can not explain cooperation in asymmetric interactions. In humans, such interactions happen often, for example humans often donate money.

@@

\textbf{Group Selection} theory posits that cooperation is favoured because of the advantage to the whole group, if selection acts at the group level in addition to the individual level. A common model for group selection work as is: the population is divided into groups. In each groups there are cooperators, which help to other group members and defectors which do not help. 
Individuals reproduce proportional to their fitness. Offspring are added to the same group.
If a group reaches a certain size it can split to two groups. A group that grow faster will split more often. Groups of cooperators are growing faster than group of defectors.
Therefore, cooperation can evolve in this model when the ratio between benefit b and cost c is more than one plus the ratio between the maximum group size n and the number of groups m:
\begin{equation} \label{groupselection}
\frac{b}{c}>1+\frac{n}{m}
\end{equation}
The group selection was criticized by biologists advocate gene-centered view of evolution. Group selection has been criticized due to the fact that the trait like cooperation evolves in the total population. According to natural selection, if cooperation took over the population it must have better fitness. However, in group selection the fitness of cooperator in the individual level is lower. The fact that a trait with a lower fitness took over the population is a contradiction. Eldar et al.~\cite{eldakar2011eight} reject this argument. They believe that this argument is a tautology and does not qualify as an argument against group selection. The distinction between individual and group selection requires a comparison of fitness differentials within and among groups in a multi group population. When a trait is evolve by group selection, despite the fact that it has lower fitness within the group, it has a better fit, all thing considered.

@@

\subsection*{Cultural Evolution of Cooperation}
Lewin-Epstein at el.~\cite{lewin2017microbes} have demonstrated that \emph{non-vertical} transmission can help to explain the evolution of cooperative behaviour. 
Non-vertical transmission could be either a horizontal or oblique. Horizontal transmission occurs between individuals from the same generation. Oblique transmission occurs from an adult to an offspring. 
Evolution under either of these transmission models is can be be more rapid than under pure vertical transmission \cite{ram2018evolution} (yr: not sure this is a good reference).

@@

Our first aim is to determine to what extent non-vertical transmission can explain the evolution of cooperation in a fully mixed population where cooperation is modelled using the prisoner's dilemma (yr: add ref).

Consider a fully mixed asexual population where individuals interact in pairs with a prisoner's dilemma payoff~\ref{fig:payoff}.

Cooperative behaviour is a dichotomous cultural trait with phenotypes $A$ and $B$ for cooperation and defection, respectively. Cooperators pay a cost $0<c<1$ and their partner gains a benefit $b$, where we assume $b>c$.
After each interaction, each individual will try to learn from its interaction partner with probability $\alpha$, and with probability $1-\alpha$ it will try to learn from random individual within the population as shown in table \ref{fig:payoff}. 
\begin{figure}[h!]
  \centering
  \includegraphics[scale=0.5]{payoff.pdf}
  \caption{Payoff Matrix}
  \label{fig:payoff}
\end{figure}

The transmission from type A will be successful with probability $T_a$ and from type B with probability $T_b$.
In the model, in each generation each individual is having n interactions and then a reproduction is occurring according to each individual fitness and phenotypes A and B are vertically transmitted from parent to offspring. In general the vertical transition can be imperfect which means that with probability $v$ offspring inherits its parent’s phenotype and with probability $1-v$ it inherits a phenotype from its parent population. After reproduction the offspring replace the parent generation.
\begin{figure}[h!]
  \centering
  \includegraphics[scale=1]{figure.pdf}
  \caption{Horizontal Transmission}
  \label{fig:figure}
\end{figure}

We define $p$ to be the frequency of newborn with phenotype A and $q=1-p$ to be the frequency of newborn with phenotype B. In addition, we define $\hat{p},\hat{q}$ to be the frequency of phenotypes A and B before reproduction and after interaction.

\begin{equation} \label{d}
\hat{p} = p^2\left[1-\left(1-\alpha\right)qT_B\right] + pq\alpha T_A + pq\alpha(1-T_B) + pq(1-\alpha)[pT_A+q(1-T_B)]
\end{equation}

The mean fitness of the population in each generation is:
\begin{equation} \label{d}
\overline{\omega} = 1 + p(b-c)
\end{equation}
As shown in eq. \eqref{vt} the probability to get a offspring with phenotype B given that its parent is of phenotype A is equal probability of imperfect vertical transmission times the proportion of A in the population. 
\begin{equation} \label{vt}
(1-v)\hat{q}
\end{equation}

Given p the frequency of phenotype A at the current generation, we are interested in frequency of A in the next generation. We have three possible interaction types A interacts with A, A interacts with B and B interacts with B.

We focus on the frequency of phenotype A. We define $x_{AA}$ to be the frequency of offspring that their parent was of phenotype A and interacted with another individual of type A. Similarly we define $x_{AB}, \, x_{BA}, \, x_{BB}$.

\begin{equation} 
\begin{split}\label{aim1model0}
 x_{AA} &= \frac{1}{\overline{\omega}}p^2(1+b-c)[(1-(1-\alpha)T_Bq)(1-(1-v)\hat{q})+(1-\alpha)T_Bq(1-v)\hat{p}]\\
 x_{AB} &= \frac{1}{\overline{\omega}}pq(1-c)[(1-(1-\alpha)T_Bq-\alpha T_B)(1-(1-v)\hat{q}) \\ &+ (\alpha T_B+(1-\alpha)T_Bq)(1-v)\hat{p}]\\
 x_{BA} &= \frac{1}{\overline{\omega}}pq(1+b)[(1-(1-\alpha)T_Ap-\alpha T_A)((1-v)\hat{p})\\ & +(\alpha T_A+(1-\alpha)T_A p)(1-(1-v)\hat{q})] \\
 x_{BB} &= \frac{1}{\overline{\omega}}q^2[(1-(1-\alpha)T_Ap)(1-v)\hat{p})+(1-\alpha)T_Ap(1-(1-v)\hat{q})]
\end{split}
\end{equation}
We define p' to be the frequency of individuals of phenotype A in the next generation.
\begin{equation} 
\begin{split}\label{aim1model}
& p' = x_{AA} + x_{AB} + x_{BA} + x_{BB}
\end{split}
\end{equation}

The question we are interested in is under what conditions the cooperative behaviour represented by phenotype A will not become extinct. In order to do that we will investigate when $p' > p$.
We can make a simplify assumption that vertical transmission is perfect, $v=1$ and then we will be able to solve the inequation more easily. Under this assumption equations \eqref{aim1model0},\eqref{aim1model} become

\begin{equation} 
\begin{split}\label{modeleq}
p' =& \frac{1}{\overline{\omega}} \{ p^2(1+b-c)[1-(1-\alpha)T_Bq] + pq(1-c)[1-(1-\alpha)T_Bq-\alpha T_B)]\\
&+ pq(1+b)[\alpha T_A+(1-\alpha)T_Ap)] + q^2(1-\alpha)T_Ap\}.
\end{split}
\end{equation}
However, we are also interested when this assumption does not apply, meaning that $v<1$. In this, We will to run numerical simulation in order to better understand under what conditions cooperative behaviour will take over the population. 
\\\textbf{\underline{Preliminary Results}}
We have preliminary result only when $v=1$.
\\We get that $p'>p$ when (see appendix A):
\begin{equation} 
\begin{split} \label{result0}
 c(1-T_B) < & \,  bqT_B(1-\alpha) - bqT_A(1-\alpha) + (T_A-T_B) + b(T_A - T_B) + \alpha bT_B
\end{split}
\end{equation}
We want to investigate few special cases:
\\\underline{Special case 1 - $T_A = T_B = T$:} then, eq. \eqref{result0} becomes:
\begin{equation}
\begin{split} \label{result}
 \frac{1-T}{\alpha T}< \frac{b}{c}
\end{split}
\end{equation}
Therefore, cooperative behaviour will take over the population when eq. \eqref{result} is satisfied.  
\\\underline{Special case 2 - $\alpha =1$:} in this case, horizontal transmission can only occur as a result of interaction.
\\Eq. \eqref{result0} becomes:
\begin{equation} 
\begin{split} \label{result3}
c(1-T_B)  < & \, T_A -T_B +b(T_A-T_B)
\end{split}
\end{equation}
Therefore, cooperative behaviour will take over the population when eq. \eqref{result3} is satisfied. 
This result is in accordance with the result presented by Lewin-Epstein at el. \cite{Ohad}. 
\\\underline{Special case 3 - $\alpha = 0$:} in this case horizontal transmission can not occur as a result of interaction.  
Then, eq. \eqref{result0} becomes:
\begin{equation} 
\begin{split} \label{result4}
0 < & \,  - bq(1-T_B)- bT_B-c(1-T_B)-T_B
\end {split}
\end{equation}
Which is always false because all of the elements are negative. Therefore, cooperation will eventually disappear from population.
\\Those are the results we have so far. However, there is a still work need to be done. \\
We will analyze the equilibrium points in this model. We know that $p=1$ is an equilibrium point because if the all individuals in the population are cooperators then in next generation all individual will be cooperator as well. Same for $p=0$. We will check if there are any more equilibrium points. For each equilibrium point we will determine whether it is stable equilibrium or not. 
\\So far, we assumed that $v=1$. We believe it will be interesting to test the effect of oblique transmission ($v\neq 1$) since oblique transmission is common in cultural evolution models. Therefore, we are planning to simulations to investigate the case when $v\neq 1$.

>>>>

\documentclass{article}
\usepackage[margin=0.5in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage[displaymath, mathlines]{lineno}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{multirow}

@@

\textbf{Kin selection}, proposing that natural selection can favour cooperative behaviour between kin. The importance of relatedness to the evolution of cooperation and altruism was shown by Hamilton~\cite{hamilton1964genetical}. According to Hamilton, kin selection causes allele to increase in frequency when the genetic relatedness $r$ of a recipient to an actor multiplied by the benefit $b$ to the recipient is greater than the reproductive cost $c$ to the actor. This is also known as Hamilton's Rule:

\begin{equation} \label{eq:hamilton_rule}
c<rb
\end{equation}

There is an ongoing debate about to what extent kin selection explains evolution of cooperation and altruism.
It has been suggested that kin selection to explain the cooperative behaviour of eusocial insects like the honey bee.
The most significant argument against kin selection is that cooperation can evolve with zero relatedness~\cite{wilson2005kin}. This makes Hamilton's rule incomplete according to Wilson~\cite{wilson2005kin}. Foster et~al.~\cite{foster2006kin} reject this claim. They argue that altruism without relatedness can not evolve. They refer us to Hamilton who claimed that relatedness can arise without recent common ancestry. 
Wilson also criticises kin selection on the grounds that environmental or ecological factors probably be more important than relatedness in determining social actions. On the other hand, Foster et~al.~\cite{foster2006kin} argue that kin selection does not ignore ecology. Hamilton’s rule shows
that environmental factors causing a high benefit-to-cost ratio will favour cooperation.

\textbf{Reciprocity} suggests repeating interactions or individual recognition as key factors for explaining the evolution of cooperation. In \textit{direct reciprocity} there are a repeated encounters between the same two individuals. In every encounter, each player has a choice between cooperation and defection. If I cooperate now, you may cooperate later. Hence, it may pay off to cooperate.
This game-theoretic framework is known as the repeated Prisoner's Dilemma. 
%A well known strategy to play this game is called  'tit-for-tat'. This strategy always starts with a cooperation, then it does whatever the other player has done in the previous round: a cooperation for a cooperation, a defection for a defection. There a unlimited number of possible strategies to play this game. However, 
Direct reciprocity can only lead to the evolution of cooperation if the cost is smaller than $w$ the probability for another encounter between the same two individuals multiplied by the benefit.

\begin{equation} \label{eq:reciprocity}
c<bw
\end{equation}

Direct reciprocity assumes that both players are in a position to cooperate. Direct reciprocity can not explain cooperation in asymmetric interactions. In humans, such interactions happen often, for example humans often donate money.

@@

\textbf{Group Selection} theory posits that cooperation is favoured because of the advantage to the whole group, if selection acts at the group level in addition to the individual level. A common model for group selection work as is: the population is divided into groups. In each groups there are cooperators, which help to other group members and defectors which do not help. 
Individuals reproduce proportional to their fitness. Offspring are added to the same group.
If a group reaches a certain size it can split to two groups. A group that grow faster will split more often. Groups of cooperators are growing faster than group of defectors.
Therefore, cooperation can evolve in this model when the ratio between benefit b and cost c is more than one plus the ratio between the maximum group size n and the number of groups m:

\begin{equation} \label{eq:groupselection}
\frac{b}{c}>1+\frac{n}{m}
\end{equation}

Group selection was criticized by biologists advocate gene-centered view of evolution. Group selection has been criticized due to the fact that the trait like cooperation evolves in the total population. According to natural selection, if cooperation took over the population it must have better fitness. However, in group selection the fitness of cooperator in the individual level is lower. The fact that a trait with a lower fitness took over the population is a contradiction. Eldar et al.~\cite{eldakar2011eight} reject this argument. They believe that this argument is a tautology and does not qualify as an argument against group selection. The distinction between individual and group selection requires a comparison of fitness differentials within and among groups in a multi group population. When a trait is evolve by group selection, despite the fact that it has lower fitness within the group, it has a better fit, all thing considered.

@@

\subsection*{Cultural Evolution of Cooperation}
Lewin-Epstein at el.~\cite{lewin2017microbes} have demonstrated that \emph{non-vertical} transmission can help to explain the evolution of cooperative behaviour. 
Non-vertical transmission could be either a horizontal or oblique. Horizontal transmission occurs between individuals from the same generation. Oblique transmission occurs from an adult to an offspring. 
Evolution under either of these transmission models is can be be more rapid than under pure vertical transmission~\cite{ram2018evolution} (yr: not sure this is a good reference).

@@

First, we focus on the evolution of cooperation in a fully mixed population where cooperation is modelled using the prisoner's dilemma (yr: add ref).

Consider a very large population whose members are characterised by their phenotype $\phi$, which can be of two types, $\phi=A$ for cooperators or $\phi=B$ for defectors, with corresponding fitness values $w_A$ and $w_B$, which depend on the frequency of the phenotypes (see below).
An offspring inherits its phenotype from its parent via vertical transmission with probability $v$ or from a random individual in the parental population via oblique transmission with probability $(1-v)$. 
Following~\cite{ram2018evolution}, given that the parent phenotype is $\phi$ and assuming uniparental inheritance (yr: cite Zefferman Behav Ecol 2016), the conditional probability that the phenotype $\phi'$ of the offspring is $A$ is

\begin{equation} \label{eq:vertical_oblique_transmission}
P(\phi'=A \mid \phi) = \begin{cases}
v + (1-v)p, & \text{if } \phi=A \\
(1-v)p, & \text{if } \phi=B
\end{cases},
\end{equation}
where $p=P(\phi=A)$ is the frequency of $A$ among adults in the parental generation.

Therefore, the frequency $\hat{p}$ of  phenotype $A$ among juveniles (after selection and vertical and oblique transmission) is

\begin{equation}\label{eq:horizontal}
\begin{aligned}
\hat{p}
& = \frac{w_A(p)}{\bar{w}} p [v + (1-v)p] + \frac{w_B(p)}{\bar{w}} (1-p) [(1-v)p] \\
& = v \frac{w_A}{\bar{w}} p + (1-v) p,
\end{aligned}
\end{equation}
where $\bar{w}=p w_A + (1-p) w_B$ is the population mean fitness.

Fitness is determined by a prisoner's dilemma dynamic. Specifically, individuals interact in pairs, cooperators pay a cost $0<c<1$, and their partner gains a benefit $b$, where we assume $b>c$. The following payoff matrix shows the fitness of an individual with phenotype $\phi_1$ when interacting with a partner with phenotype $\phi_2$ ($b>c>0$):

\begin{table}[h]\label{table:prisoner_payoff}
\centering
\begin{tabular}{lll}
           & $\phi_2=A$ & $\phi_2=B$ \\
$\phi_1=A$ & $b-c$      & $-c$       \\
$\phi_1=B$ & $b$        & $0$       
\end{tabular}
\end{table}

Interactions occur randomly.
So, two individuals with phenotype $A$ interact with probability $\hat{p}^2$, two individuals with phenotype $B$ interact with probability $(1-\hat{p})^2$, and two individuals with different phenotypes interact with probability $2\hat{p}(1-\hat{p})$. 
Therefore, the fitness values of the two phenotypes, $w_A$ and $w_B$ are

\begin{equation}\label{eq:fitness}
\begin{aligned}
w_A(\hat{p}) & = \hat{p}(1+b-c) + (1-\hat{p})(1-c) = 1 - c + \hat{p} b \\
w_B(\hat{p}) & = \hat{p}(1+b) + (1-\hat{p}) = 1 + \hat{p} b,
\end{aligned}
\end{equation}
and the population mean fitness is therefore
$\bar{w} = \hat{p} w_A + (1-\hat{p}) w_B = 1 + \hat{p}(b-c)$.

Horizontal cultural transmission, which occurs between peers, may occur between interaction partners with probability $\alpha$, or between a random pair with probability $1-\alpha$ (see~\textbf{Fig.~\ref{fig:horizontal}}).
Horizontal transmission is not always successful, as one peer may reject the other's phenotype. The probability for successful transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively.

\begin{figure}[h!]
  \centering
  \includegraphics[scale=1]{figure.pdf}
  \caption{\textbf{Cultural horizontal transmission.} Transmission occurs between interaction partners with probability $\alpha$ (left) or between two random peers with probability $1-\alpha$.}
  \label{fig:horizontal}
\end{figure}

Therefore, the next-generation frequency $p'$ of adults with phenotype $A$ after interaction and horizontal transmission is (yr: please check)
\begin{equation}\label{eq:horizontal}
\begin{aligned}
p'
& = \hat{p}^2 [\alpha + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + \hat{p}(1-\hat{p}) [\alpha(1-T_b) + (1-\alpha)(\hat{p} + (1-\hat{p})(1-T_B))] \\
& + (1-\hat{p})\hat{p} [\alpha T_A + (1-\alpha) \hat{p} T_A ] \\
& + (1-\hat{p})^2 [(1-\alpha) \hat{p} T_A]
\end{aligned}
\end{equation}

\section*{Results}

We start by inspecting specific cases, for which we can derive general results. 
Afterwards, we use numerical simulation to analyse more complex cases.

\paragraph*{Without oblique transmission, $v=1$.} 
With only vertical transmission eq.~\ref{eq:vertical_oblique_transmission} becomes
\begin{equation}
\hat{p} = \frac{w_A}{\bar{w}} p,
\end{equation}
and the change in frequency of adults with phenotype $A$ can be written as

\begin{equation} 
\begin{split}\label{eq:model_v=1}
p' =& \frac{1}{\overline{\omega}} \{ p^2(1+b-c)[1-(1-\alpha)T_Bq] + pq(1-c)[1-(1-\alpha)T_Bq-\alpha T_B)]\\
&+ pq(1+b)[\alpha T_A+(1-\alpha)T_Ap)] + q^2(1-\alpha)T_Ap\}.
\end{split}
\end{equation}
However, we are also interested when this assumption does not apply, meaning that $v<1$. In this, We will to run numerical simulation in order to better understand under what conditions cooperative behaviour will take over the population. 
\\\textbf{\underline{Preliminary Results}}
We have preliminary result only when $v=1$.
\\We get that $p'>p$ when (see appendix A):
\begin{equation} 
\begin{split} \label{result0}
 c(1-T_B) < & \,  bqT_B(1-\alpha) - bqT_A(1-\alpha) + (T_A-T_B) + b(T_A - T_B) + \alpha bT_B
\end{split}
\end{equation}
We want to investigate few special cases:
\\\underline{Special case 1 - $T_A = T_B = T$:} then, eq. \eqref{result0} becomes:
\begin{equation}
\begin{split} \label{result}
 \frac{1-T}{\alpha T}< \frac{b}{c}
\end{split}
\end{equation}
Therefore, cooperative behaviour will take over the population when eq. \eqref{result} is satisfied.  
\\\underline{Special case 2 - $\alpha =1$:} in this case, horizontal transmission can only occur as a result of interaction.
\\Eq. \eqref{result0} becomes:
\begin{equation} 
\begin{split} \label{result3}
c(1-T_B)  < & \, T_A -T_B +b(T_A-T_B)
\end{split}
\end{equation}
Therefore, cooperative behaviour will take over the population when eq. \eqref{result3} is satisfied. 
This result is in accordance with the result presented by Lewin-Epstein at el. \cite{Ohad}. 
\\\underline{Special case 3 - $\alpha = 0$:} in this case horizontal transmission can not occur as a result of interaction.  
Then, eq. \eqref{result0} becomes:
\begin{equation} 
\begin{split} \label{result4}
0 < & \,  - bq(1-T_B)- bT_B-c(1-T_B)-T_B
\end {split}
\end{equation}
Which is always false because all of the elements are negative. Therefore, cooperation will eventually disappear from population.
\\Those are the results we have so far. However, there is a still work need to be done. \\
We will analyze the equilibrium points in this model. We know that $p=1$ is an equilibrium point because if the all individuals in the population are cooperators then in next generation all individual will be cooperator as well. Same for $p=0$. We will check if there are any more equilibrium points. For each equilibrium point we will determine whether it is stable equilibrium or not. 
\\So far, we assumed that $v=1$. We believe it will be interesting to test the effect of oblique transmission ($v\neq 1$) since oblique transmission is common in cultural evolution models. Therefore, we are planning to simulations to investigate the case when $v\neq 1$.
